Relationships in the extremes and their influence on competition and coexistence

### 2021-09-17 paper/Paper.Rnw
parahraph 4 of intro
topic sentence for paragraph 5

%***DAN: Jasmin, there are a lot of disjointed sentences in the below paragraph, at least they appear disjointed to me.
%Can you please say in a comment right below this one what the topic sentence is? As usual, if you are having trouble 
%determining what the topic sentence it, it may mean you don't have one, and you need to determine what you're doing 
%with this paragraph and make a topic sentence. 
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and extent \citep{Meehl_2004, Ummenhofer_2017}. 
The rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. are a product of climate change and have implications for coexistence and competition. Changes in means can be attributed to extreme events; however changes in means are more-so a result of gradual increases or decreases over time in an environmental variable. 
Extreme events push environmental variables to their extremes, increasing variance. Effects on competition and coexistence due to climate change are more widely researched through studying changes in the mean of environmental variables. 
For example, decreased growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversely, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease.    
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
Extreme events can have important ecological impacts for controphic 
%***DAN: I don't know what "controphic" means and when I google it I get nothing. Perhaps you intended a different word?
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in gerneral (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

>>>>

%***DAN: Jasmin, there are a lot of disjointed sentences in the below paragraph, at least they appear disjointed to me.
%Can you please say in a comment right below this one what the topic sentence is? As usual, if you are having trouble 
%determining what the topic sentence it, it may mean you don't have one, and you need to determine what you're doing 
%with this paragraph and make a topic sentence. 
%***JAS: I am trying to relay the message that climate change and extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and extent \citep{Meehl_2004, Ummenhofer_2017}. 
Thus, the rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. implications for coexistence and competition. Changes in means can be attributed to extreme events; however changes in means are more-so a result of gradual increases or decreases over time in an environmental variable. 
Extreme events push environmental variables to their extremes, increasing variance. Effects on competition and coexistence due to climate change are more widely researched through studying changes in the mean of environmental variables. 
For example, decreased growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversely, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease.    
In contrast, increased environmental variance has been less commonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Moreover, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
%**JAS: I intended for the first sentence to be the topic sentence 
% a better one:
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Extreme events can have important ecological impacts for
%***DAN: I don't know what "controphic" means and when I google it I get nothing. Perhaps you intended a different word?
% can cut out 
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).
